---
title: "Gene Expression Analysis" 
author: "Author: Your Name"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
        collapsed: true
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    code_folding: show
    number_sections: false
fontsize: 14pt
bibliography: bibtex.bib
type: docs
weight: 304
---

<!---
- Compile from command-line
Rscript -e "rmarkdown::render('exprAnalyisDemo.Rmd', c('html_document'), clean=FALSE)"
-->

# Initialize project 

The instructions will clone the pregenerated code for this exercise from GitHub into an R project on RStudio Cloud.

1. Create a free personal account on RStudio Cloud [here](https://rstudio.cloud/)
2. Go to RStudio Cloud [here](https://rstudio.cloud/projects/)
3. Select under blue _New Project_ menu: `New Project from Git Repository`
4. Provide this URL for BIOL119 repos: https://github.com/tgirke/BIOL119.git 
5. Optional: to work with your own GitHub repos and push changes to it, 
   then it will be necessary to set your GitHub identity with:
   - git config --global user.email "..."
   - git config --global user.name "..."

# Install required packages
```{r package_installs, eval=FALSE, message=FALSE, warning=FALSE}
if(!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c("GEOquery", "limma", "clusterProfiler", "org.Hs.eg.db", "enrichplot", "pheatmap", "RColorBrewer"))
```

# Import data from GEO

The following uses the `GEOquery` package to automatically download from GEO 
a expression data set that involves a chemical treatment. The specific data set chosen for this exercise is experiment [GDS2778](https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS2778). 
This Affymetrix gene expression profiling experiment from Gillis _et al._ [-@Gillis2007-qk] aims to study the effect of the air pollutant 
1,2,4-benzenetriol on blood cells to provide insight into the molecular basis of benzene cytotoxicity. 
More detailed annotation information about this experiment is available under Reference 
Series [GSE7664 URL](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE7664). 

```{r import_data_from_geo, eval=TRUE, message=FALSE}
library(GEOquery) 
gds <- getGEO("GDS2778") # Download GDS2778 data from GEO
eset <- GDS2eSet(gds)
pData(eset)[ ,1:2] # Inspect experimental design 
exprs(eset)[1:4, ] # Inspect expression data that are RMA normalized and on log2 scale
```

# DEG analysis with Limma

Text and citation... [@Ritchie2015-rl]

```{r deg_analysis, eval=TRUE, message=FALSE}
library(limma) # Loads limma library.
targets <- pData(eset)
design <- model.matrix(~ -1+factor(as.character(targets$agent)))
colnames(design) <- c("Treatment", "Control") 
fit <- lmFit(eset, design) # Fits a linear model for each gene based on the given series of arrays.
contrast.matrix <- makeContrasts(Control-Treatment, levels=design)
fit2 <- contrasts.fit(fit, contrast.matrix) # Computes estimated coefficients and standard errors for a given set of contrasts.
fit2 <- eBayes(fit2)
deg_df <- topTable(fit2, coef=1, adjust="fdr", sort.by="B", number=Inf)
deg_df <- deg_df[!is.na(deg_df$adj.P.Val),]
```

# Over-represention analysis (ORA) with clusterProfiler

text and citation ... [@Yu2012-hr]

```{r overrepresentation_analysis, eval=TRUE, message=FALSE}
cutoff <- 0.05 # Cutoff to use for filtering on adjusted p-value (FDR)
ids <- deg_df[deg_df$adj.P.Val <= cutoff, "Gene.ID"]
ids <- ids[!grepl("/", ids)] # Removes ambiguous mappings
ids <- ids[grepl("", ids)]
library(clusterProfiler); library(org.Hs.eg.db); library(enrichplot)
ego_mf <- enrichGO(gene=ids, OrgDb=org.Hs.eg.db, ont="MF", pAdjustMethod="BH", pvalueCutoff=0.1, readable=TRUE)
ego_bp <- enrichGO(gene=ids, OrgDb=org.Hs.eg.db, ont="BP", pAdjustMethod="BH", pvalueCutoff=0.1, readable=TRUE)
```

## Visualize BP and MF ontology results
```{r plot_enrichment, eval=TRUE, message=FALSE}
dim(ego_mf)
head(ego_mf)
barplot(ego_mf, showCategory=10)
goplot(ego_mf)
dim(ego_bp)
head(ego_bp)
barplot(ego_bp, showCategory=10)
```

# Clustering

...

```{r clustering, eval=TRUE, message=FALSE}
library(pheatmap); library("RColorBrewer")
cutoff <- 0.05 # Cutoff to use for filtering on adjusted p-value (FDR)
affy_ids <- row.names(deg_df[deg_df$adj.P.Val <= cutoff, ])
deg_ma <- exprs(eset)[affy_ids, ] 
pheatmap(deg_ma, scale="row", color=brewer.pal(9, "Blues"))
```

## Session Info

```{r sessionInfo}
sessionInfo()
```

# References
